---
title: "Handling Missing Data: How R Does It Differently Than SAS"
date: "2026-02-08"
author: "Harsh Verma"
categories: [SAS-to-R, Data Cleaning, Clinical]
description: "Why R’s NA is completely different from SAS's period (.) and how missing data propagates."
---

## The Missing Data Shock

If there's one thing that initially frustrates SAS programmers transitioning to R, it's missing data. 

In SAS, missing numeric data is denoted by a period (`.`) and missing character data is denoted by a blank (`""`). It’s simple, it's consistent across datasets, and SAS procedures are built to politely ignore them.

```sas
/* SAS Code */
data example;
  x = 10;
  y = .;
  total = sum(x, y); /* total evaluates to 10 */
  mean_val = mean(x, y); /* mean evaluates to 10 */
run;
```
If you run this in SAS, `total` is 10 and `mean_val` is 10. SAS assumes you want it to ignore the missing value and proceed with what it has.

## Welcome to R: The Infection of NA

In R, missing data is denoted by `NA` (Not Available). And unlike SAS, R operates under the philosophy that **missing data is infectious**. If you don’t know part of an equation, you logically cannot know the result.

```{r}
# R Code
x <- 10
y <- NA

# Basic arithmetic
total <- sum(x, y) 
mean_val <- mean(c(x, y))

print(total)
# [1] NA
print(mean_val)
# [1] NA
```

R refuses to guess. If you ask it for the sum of 10 and an unknown number, it sensibly concludes the answer is unknown.

### The Fix: `na.rm = TRUE`

To make R behave somewhat like SAS, you simply have to instruct it to "remove NAs" before performing an operation. This is done with the `na.rm = TRUE` argument inside functions like `sum()`, `mean()`, and `sd()`.

```{r}
# R Code
total_rm      <- sum(x, y, na.rm = TRUE)
mean_val_rm   <- mean(c(x, y), na.rm = TRUE)

print(total_rm)
# [1] 10
print(mean_val_rm)
# [1] 10
```

This is incredibly important when summarizing clinical data. If one patient out of 500 is missing a baseline lab value, your overall baseline mean calculation will fail (`NA`) unless you explicitly set `na.rm = TRUE`.

## The Great Equality Trap

Here is another massive gotcha for SAS programmers. 

In SAS, if you want to find all rows where `AGE` is missing, you do this:
```sas
/* SAS Code */
data missing_age;
  set adsl;
  if age = .;
run;
```

If you try to perfectly translate that to R, you will crash and burn.

```{r}
# Bad R Code - Do not do this
# adsl %>% filter(age == NA)
```

Why doesn't this work? In R, the expression `NA == NA` evaluates to `NA`. You cannot check if something is equal to "unknown." So comparing `age == NA` doesn't return `TRUE` or `FALSE`; it returns a vector of `NA`s, which causes `filter()` to fail.

To check for missingness in R, you **must** use the `is.na()` function.

```{r}
# Mock ADSL Data
library(dplyr)
adsl <- data.frame(
  usubjid = 1:5,
  age = c(45, NA, 60, NA, 30),
  status = c("Completed", "Dropout", "Completed", "Randomized", "Dropout"),
  weight = c(80, 75, 90, 85, 78)
)

# Correct R Code
# To find missing values
# missing_age <- adsl %>% filter(is.na(age))

# To drop missing values
# complete_age <- adsl %>% filter(!is.na(age))
# or using tidyr
# complete_age <- adsl %>% drop_na(age)
```

## Creating Missing Data

In SAS, you might conditionally wipe out a value like this:

```sas
/* SAS Code */
if (status = 'Dropout') then weight = .;
```

In R, we typically use `{dplyr}`'s `mutate()` and `ifelse()` or `if_else()` or `case_when()`. 

```{r}
# R Code
# adsl <- adsl %>%
#  mutate(weight = if_else(status == "Dropout", NA_real_, weight))
```

Wait, what is `NA_real_`?

R is a strictly typed language. Under the hood, standard `NA` is a logical vector. When you are replacing numeric values (reals), `{dplyr}` functions like `if_else()` (which are stricter than base R's `ifelse()`) require the replacement value to be of the exact same type as the original column. Thus, you use `NA_real_` for double precision floating-point numbers, `NA_integer_` for whole numbers, and `NA_character_` for text. 

## Final Thoughts

R's strict handling of missing data forces you to be deliberate. In clinical reporting, this is a massive advantage! You want to be violently thrown an error if you try to average a column containing hidden missing data, because quietly dropping it could lead to incorrect efficacy results. 

In the next post, we'll cover how to replicate SAS's mighty `PROC FORMAT`.
