---
title: "Translating PROC FREQ and PROC MEANS into R: A Practical Guide"
date: "2026-02-05"
author: "Harsh Verma"
categories: [SAS-to-R, Tidyverse, Summary Statistics, Clinical]
description: "How to summarize your clinical data without relying on PROC FREQ or PROC MEANS."
---

## The Bread and Butter of Clinical Reporting

If you write SAS code for clinical trials, you likely type `PROC FREQ` and `PROC MEANS` in your sleep. These two procedures form the backbone of exploratory data analysis, data checking, and baseline characteristic tables.

When migrating to R, SAS programmers inevitably ask: *"What's the R equivalent of PROC MEANS?"* 

The good news? The R ecosystem (specifically the `{tidyverse}`) provides tools that are just as powerful, but vastly more flexible for integration into downstream operations. 

Let's break down how to handle tabulations and summary statistics in R.

## Replacing PROC FREQ

Let’s look at a standard frequency table for adverse event severity (AETOXGR).

```sas
/* SAS Code: PROC FREQ */
proc freq data=adae;
  tables aetoxgr / missing;
run;
```

### The Base R Way

In Base R, you have the `table()` function, which is functional but lacks the modern structure we usually want.

```{r}
# Mock ADAE Data
adae <- data.frame(
  subjid = 1:50,
  trta = sample(c("Placebo", "Active"), 50, replace = TRUE),
  aetoxgr = sample(c("MILD", "MODERATE", "SEVERE"), 50, replace = TRUE)
)

# Base R
table(adae$aetoxgr, useNA = "always")
```

Using `{dplyr}`, the replacement is the `count()` function. It’s elegant and returns a data frame (which SAS users appreciate, given that `PROC FREQ` requires an `ods output` statement to capture the results!).

```{r}
#| label: count-example
#| message: false
library(dplyr)

# dplyr equivalent
adae %>%
  count(aetoxgr)
```

Need percentages like SAS provides natively? Combine `count()` with a simple `mutate()`:

```{r}
adae %>%
  count(aetoxgr) %>%
  mutate(percent = (n / sum(n)) * 100)
```

### Cross-Tabulations

SAS:
```sas
proc freq data=adae;
  tables trta * aetoxgr;
run;
```

R `{dplyr}` + `{tidyr}`:
```{r}
library(tidyr)

adae %>%
  count(trta, aetoxgr) %>%
  pivot_wider(names_from = aetoxgr, values_from = n, values_fill = 0)
```
Notice how `pivot_wider()` reshapes the data frame exactly the way you need it for reporting—no `PROC TRANSPOSE` required!

## Replacing PROC MEANS

`PROC MEANS` (or its twin, `PROC SUMMARY`) is used to calculate N, mean, standard deviation, median, min, and max.

```sas
/* SAS Code: PROC MEANS */
proc means data=advs n mean std median min max;
  class paramcd;
  var aval;
run;
```

### The Tidyverse Way: `group_by()` + `summarize()`

In R, calculating summary statistics is a two-step pipeline: group the data, then summarize it. 

```{r}
# Mock ADVS Data
advs <- data.frame(
  subjid = rep(1:10, each = 2),
  paramcd = rep(c("SYSBP", "DIABP"), times = 10),
  aval = c(rnorm(10, 120, 10), rnorm(10, 80, 5))
)

# dplyr equivalent
advs %>%
  group_by(paramcd) %>%
  summarize(
    n = n(),
    mean_val = mean(aval, na.rm = TRUE),
    std_dev = sd(aval, na.rm = TRUE),
    median_val = median(aval, na.rm = TRUE),
    min_val = min(aval, na.rm = TRUE),
    max_val = max(aval, na.rm = TRUE)
  )
```

**Two key things to note here for SAS transplants:**
1. R is unforgiving with missing data. In SAS, `mean(1, 2, .)` is `1.5`. In R, `mean(c(1, 2, NA))` is `NA`. You almost always need to add `na.rm = TRUE` to your statistical functions to skip missing values explicitly.
2. The result is inherently a tidy, structured dataset. There's no separate "procedure output window" vs "output dataset." You just assign the result directly: `summary_stats <- advs %>% summarize(...)`

## Summarizing Multiple Variables at Once 

What if you want the mean of multiple variables? SAS makes this easy by listing them in the `var` statement. 

In R, `{dplyr}` provides the incredibly powerful `across()` function. 

```{r}
# R: Summarizing multiple numeric variables
advs %>%
  group_by(paramcd) %>%
  summarize(
    across(
      .cols = where(is.numeric), 
      .fns = list(
        mean = ~mean(.x, na.rm = TRUE),
        sd = ~sd(.x, na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  )
```
This single block of code takes every numeric column in your dataset, calculates the mean and standard deviation, and automatically names the new columns (e.g., `aval_mean`, `aval_sd`). It is essentially a programmatic macro built right into the syntax.

## Conclusion

Letting go of `PROC FREQ` and `PROC MEANS` feels strange at first. But once you realize that R functions like `count()` and `summarize()` instantly return structured data frames instead of static printouts, you'll wonder how you ever lived without them. 

Next up: The great missing data mismatch. We’ll explore why R’s `NA` is a totally different beast than SAS's period (`.`).
