---
title: "Safety TFLs: Visualizing Lab Shifts and Vital Sign Trends in R"
date: "2026-03-15"
author: "Harsh Verma"
categories: [Figures, Safety, ggplot2, Clinical]
description: "Move beyond tables: Using {ggplot2} to create powerful visual representations of laboratory shift tables and vital sign trends."
---

## The Limitations of Tables

We spend hours making Laboratory Shift Tables—those massive matrices showing how many subjects went from "Low" at baseline to "High" at Week 12. 

The problem? Most medical reviewers glance at these 100-page tables and immediately look for a figure instead. Large cross-tabulations are terrible at conveying *magnitude*. Yes, 5 patients shifted to "High" ALT, but did they barely cross the threshold, or did their liver enzymes spike 10x the Upper Limit of Normal (ULN)?

In SAS, creating complex interactive or highly customized graphics (`PROC SGPLOT` / `GTL`) is cumbersome. In R, we have the gold standard: `{ggplot2}`.

## Building a Spaghetti Plot for Vitals

Let's map out how a SAS programmer would build a trend plot of Systolic Blood Pressure (`SBP`) over time for individual patients, overlaid with a mean trend line.

```{r}
#| label: spaghetti-vitals
#| message: false
library(dplyr)
library(ggplot2)

# Simulated Vitals Data
set.seed(123)
advs <- data.frame(
  USUBJID = rep(c("101", "102", "103", "104", "105"), each = 4),
  TRTA = rep(c("Placebo", "Active", "Placebo", "Active", "Active"), each = 4),
  AVISITN = rep(c(0, 4, 8, 12), times = 5),
  AVISIT = rep(c("Baseline", "Week 4", "Week 8", "Week 12"), times = 5)
) %>%
  mutate(AVAL = round(rnorm(n(), mean = 120, sd = 15))) # Random SBP values

# Ensure AVISIT is ordered correctly as a factor!
advs$AVISIT <- factor(advs$AVISIT, levels = c("Baseline", "Week 4", "Week 8", "Week 12"))

```

### The `{ggplot2}` Pipeline

In `{ggplot2}`, we define the dataset, map the axes (aesthetics or `aes()`), and then add geometric layers (`geom_line()`, `geom_point()`).

```{r}
# Basic Spaghetti Plot
ggplot(advs, aes(x = AVISIT, y = AVAL, group = USUBJID, color = TRTA)) +
  geom_line(alpha = 0.5) +  # Faded lines for individual subjects
  geom_point(alpha = 0.5) + # Faded points
  
  # Overlay a LOESS smoothing curve (overall trend) grouping by TRTA
  geom_smooth(aes(group = TRTA), method = "loess", se = FALSE, size = 1.5) +
  
  # Clinical Formatting
  theme_minimal() +
  labs(
    title = "Systolic Blood Pressure Trends by Treatment Arm",
    x = "Visit",
    y = "Systolic Blood Pressure (mmHg)",
    color = "Treatment Arm"
  ) +
  scale_color_manual(values = c("Active" = "#0072B2", "Placebo" = "#D55E00"))
```

In a single block of code, we created individual patient trajectories while automatically calculating and overlaying a non-parametric smoothing line (`geom_smooth`) representing the treatment arm average. 

## Visualizing Lab Shifts: The Boxplot Approach

Shift tables lack magnitude. A better approach is to plot the actual lab values at Baseline vs. Maximum Post-Baseline grouped by their Baseline Status.

Let's say we have derived baseline grades for Bilirubin (`BILI`). We want to see the distribution of maximum post-baseline values.

```{r}
# Simulated Lab Data
adlb_max <- data.frame(
  USUBJID = 1:50,
  TRTA = sample(c("Placebo", "Active"), 50, replace = TRUE),
  BBASE_GRADE = sample(c("Normal", "High"), 50, replace = TRUE, prob = c(0.8, 0.2)),
  AVAL = rnorm(50, mean = 1.5, sd = 0.5) # Max Post-Baseline values
)

ggplot(adlb_max, aes(x = BBASE_GRADE, y = AVAL, fill = TRTA)) +
  geom_boxplot(outlier.color = "red", outlier.size = 2) +
  
  # Add a horizontal line representing the Upper Limit of Normal (e.g., 1.2)
  geom_hline(yintercept = 1.2, linetype = "dashed", color = "black") +
  
  annotate("text", x = 1, y = 1.25, label = "ULN Threshold", size = 3, hjust = 0) +
  
  theme_bw() +
  labs(
    title = "Max Post-Baseline Bilirubin by Baseline Grade",
    subtitle = "Red points indicate significant outliers",
    x = "Baseline Grade",
    y = "Maximum Post-Baseline Bilirubin (mg/dL)",
    fill = "Treatment Arm"
  )
```

### Why this beats a table

If a reviewer looks at this plot, they can instantly see:
1. Did the "Normal" baseline subjects trend higher in the Active arm? (Look at the median line of the boxplot).
2. Are there any catastrophic outliers? (The red dots).
3. How many people crossed the ULN threshold who were previously normal? (The distribution above the dashed line).

## Reproducibility Through Code

The greatest benefit of generating figures in R is reproducibility. You can wrap these `ggplot` objects into generalized functions, stick them into an internal company R package (`e.g., KnowledgeCentReggplots`), and instantly generate stylized, branded, SAP-approved figures for every study with a single line of code.

Next up, we dive into Efficacy! We will tackle Survival Analysis and the legendary Kaplan-Meier plot.
