---
title: "Replicating SAS Formats and Informats in R (Using Factors and Lookups)"
date: "2026-02-12"
author: "Harsh Verma"
categories: [SAS-to-R, Categorical Data, Clinical]
description: "How to translate PROC FORMAT logic into R using `{forcats}`, named vectors, and joins."
---

## Life Without PROC FORMAT

There are few features SAS programmers love more than `PROC FORMAT`. It allows you to store the underlying raw data (like `1`, `2`, `3`) but display beautiful, human-readable labels (`"Mild"`, `"Moderate"`, `"Severe"`) in all your outputs, without modifying the actual dataset. It's clean, efficient, and essential for clinical TFLs.

When attempting to do this in R, SAS programmers often try to permanently overwrite the column string with nested conditional logic. 
*"Wait, do I actually have to replace my 1s with 'Mild' in the dataframe?"*

The answer is: Not necessarily! R has robust systems for handling categorical data and lookups that mimic what `PROC FORMAT` does, but they work quite differently.

## Method 1: Factors (The Closest Native Equivalent)

In R, a **factor** is used to store categorical data. It is literally an integer vector with an associated set of character "labels" (or levels). This makes it conceptually very similar to applying a format.

```sas
/* SAS Code */
proc format;
  value sev
    1 = "Mild"
    2 = "Moderate"
    3 = "Severe";
run;

data adae;
  set adae;
  format aesev sev.;
run;
```

In R, we create factors either using base R `factor()` or the `{forcats}` package (part of the tidyverse).

```{r}
# R Code
library(dplyr)
library(forcats)

# Sample data
adae <- data.frame(subjid = 1:5, aesev_n = c(1, 3, 2, 2, 1))

# Creating a factor
adae <- adae %>%
  mutate(
    aesev = factor(
      aesev_n,
      levels = c(1, 2, 3),
      labels = c("Mild", "Moderate", "Severe")
    )
  )

print(adae)
```

**Why Factors are Powerful for Clinical Work:**
1. **Sorting:** By default, R sorts characters alphabetically (Mild, Moderate, Severe). But 'Mild' shouldn't come after 'Severe'. Factors remember the underlying integer order you define in `levels`. This is critical for TFLs!
2. **Missing levels:** If an adverse event dataset doesn't contain any "Severe" events, `table(adae$aesev)` in R will still show a count of 0 for Severe because R knows it’s a valid possible level. This ensures your tables don't suddenly drop columns!

## Method 2: Lookup Tables (Joins)

If you are mapping complex text, ICD-10 codes, or MedDRA dictionaries, you shouldn't hard-code factors. Instead, you create a lookup table (a data frame) and join it. This is similar to creating a format from a dataset in SAS (`cntlin`).

```sas
/* SAS Code */
proc sql;
  create table adae_mapped as
  select a.*, f.term_text
  from adae a
  left join meddra_dict f
  on a.pt_code = f.pt_code;
quit;
```

In R, we use `{dplyr}`'s `left_join()`.

```{r}
# R Code
# Dictionary dataset
meddra_dict <- data.frame(
  pt_code = c(1001, 1002, 1003),
  term_text = c("Headache", "Nausea", "Fatigue")
)

# Patient data
adae <- data.frame(subjid = 1:3, pt_code = c(1002, 1001, 1002))

# Applying the format via left join
adae_mapped <- adae %>%
  left_join(meddra_dict, by = "pt_code")

print(adae_mapped)
```

This is the standard approach for merging in decodes (e.g., merging `PARAMCD` and `PARAM`).

## Method 3: Named Vectors (The Swiss Army Knife)

Sometimes you just want a quick inline format replacement without building a whole data frame. In SAS, you might use a messy `if/else if/else` block. In R, you can use a **named vector**. 

A named vector is like a tiny dictionary where the "names" are your raw values, and the "values" are your formatted labels. You can use R's subsetting `[` operator to replace values quickly.

```{r}
# Define our "format"
fmt_sex <- c("M" = "Male", "F" = "Female", "U" = "Unknown")

adsl <- data.frame(subjid = 1:4, sex = c("M", "F", "F", "U"))

# Apply the format by using the raw column to subset the named vector
adsl$sex_decoded <- unname(fmt_sex[adsl$sex])

print(adsl)
```

Alternatively, in a `#dplyr` pipeline, you can use `recode()` or `case_match()`.

```{r}
adsl <- adsl %>%
  mutate(
    sex_decoded = case_match(sex, 
      "M" ~ "Male",
      "F" ~ "Female",
      "U" ~ "Unknown"
    )
  )
```

## Wrapping Up

R doesn't give you a direct 1-to-1 substitute for `PROC FORMAT`. Instead, it forces you to think about what you actually want to achieve:
* **Need distinct ordering for tables?** Use `factor()`.
* **Need to merge a dictionary of thousands of terms?** Use `left_join()`.
* **Need a quick, hard-coded string replacement?** Use a named vector or `case_match()`.

You are now well-equipped to handle the primary hurdles of moving your data operations from SAS to R. Up next: Building SDTMs directly in R!
