---
title: "Efficacy TFLs: Survival Analysis and Kaplan-Meier Plots using {survival} and {survminer}"
date: "2026-03-18"
author: "Harsh Verma"
categories: [Efficacy, Survival, ggplot2, Clinical]
description: "How to fit Cox Proportional Hazard models and draw beautiful Kaplan-Meier curves natively in R without SAS LIFETEST."
---

## The Efficacy Endpoint: Time-to-Event

In oncology trials, the primary efficacy endpoint is often a time-to-event variable, such as Overall Survival (OS) or Progression-Free Survival (PFS).

In SAS, analyzing these endpoints generally requires two heavy-hitting procedures: `PROC LIFETEST` (for generating Kaplan-Meier curves and Log-Rank tests) and `PROC PHREG` (for Cox Proportional Hazards modeling).

In R, the standard isn't just "good enough"—it's the industry standard across all academic statistics. We use the `{survival}` package, authored originally by Terry Therneau (who literally wrote the book on modeling survival data). We pair it with `{survminer}` for publication-ready graphics.

## Fitting the Model

Let's assume our `ADSL` dataset contains `AVAL` (time in months to event) and `CNSR` (censor flag). Note: In `{survival}`, the censor indicator needs to be formulated as *Event = 1 (Dead), Censor = 0 (Alive)*, whereas SAS sometimes defaults to the opposite.

```{r}
#| label: survival-analysis
#| message: false
library(dplyr)
library(survival)

# Simulated ADTTE (Time to Event) dataset
adtte <- data.frame(
  USUBJID = 1:200,
  TRTA = sample(c("Placebo", "Active"), 200, replace = TRUE),
  AVAL = rexp(200, rate = 0.1), # Simulated survival time
  CNSR = sample(c(0, 1), 200, replace = TRUE, prob = c(0.7, 0.3)) # 0 = event, 1 = censored
)

# Convert SAS CNSR (1=Censored, 0=Event) to R Status (0=Censored, 1=Event)
adtte <- adtte %>%
  mutate(STATUS = if_else(CNSR == 0, 1, 0))

# 1. Create the Survival Object
surv_obj <- Surv(time = adtte$AVAL, event = adtte$STATUS)

# 2. Fit the Kaplan-Meier model
km_fit <- survfit(surv_obj ~ TRTA, data = adtte)

# 3. Fit a Cox Proportional Hazards Model (for Hazard Ratio)
cox_fit <- coxph(surv_obj ~ TRTA, data = adtte)

# Output summary
summary(cox_fit)
```

The output of `summary(cox_fit)` provides the Hazard Ratio (`exp(coef)`), the confidence intervals, and the p-values immediately. You can easily extract these using the `{broom}` package to populate your Table 14.2.1.X.

## Visualizing the Kaplan-Meier Curve

Now, for the holy grail of oncology TFLs: The Kaplan-Meier Plot.

In SAS, customizing `PROC SGPLOT` to include a risk table at the bottom, adjusting the censor tick marks, and formatting the p-value text is notoriously finicky.

With `{survminer}` in R, it is handled automatically through a clean `ggplot2` wrapper.

```{r}
#| label: km-plot
#| message: false
library(survminer)

ggsurvplot(
  fit = km_fit,
  data = adtte,
  
  # Styling the Lines
  palette = c("#0072B2", "#D55E00"), # Colorblind friendly
  size = 1.2,
  linetype = "solid",
  
  # Censor marks
  censor = TRUE,
  censor.shape = 124, # Represents a vertical tick
  censor.size = 3,
  
  # Axes & Grid
  xlim = c(0, max(adtte$AVAL)),
  break.time.by = 5,
  xlab = "Time in Months",
  ylab = "Probability of Survival",
  
  # Confidence Intervals 
  conf.int = TRUE,
  
  # Risk Table at the bottom
  risk.table = TRUE,
  risk.table.height = 0.25,
  risk.table.title = "Number at risk",
  
  # Statistical reporting
  pval = TRUE, # Automatically calculates and prints Log-rank p-value
  pval.coord = c(2, 0.2)
)
```

### Why this is highly superior for Clinical Trials:
1. **The Risk Table:** Adding `risk.table = TRUE` automatically generates an aligned table beneath the plot showing subjects at risk across the time bins.
2. **Dynamic P-Values:** It inherently understands the fitted model (`km_fit`) and calculates the test statistic dynamically, placing it on the chart.
3. **Themes:** It inherits `{ggplot2}` theming, meaning your company can define a `theme_company_km()` and apply it globally to all charts.

Next up, we will tackle the visually stunning (and often structurally complex) **Waterfall and Spider Plots** for solid tumor response (RECIST) criteria!
