---
title: "Efficacy TFLs: Creating Waterfall and Spider Plots for Tumor Response"
date: "2026-03-22"
author: "Harsh Verma"
categories: [Efficacy, Oncology, ggplot2, Visualizations]
description: "A step-by-step guide to plotting RECIST tumor responses using Waterfall and Spider plots in R."
---

## Oncology Visualizations 

Survival curves tell us *when* a patient failed therapy. But in early-phase solid tumor oncology trials (measuring RECIST criteria), we are deeply interested in *how much* their tumors grew or shrank over time relative to baseline.

Two plots dominated this space: The **Waterfall Plot** (Best Overall Response per patient) and the **Spider Plot** (Tumor burden trajectory over time).

Getting these right in SAS `PROC SGPLOT` requires massive data manipulation to generate sorted dummy variables for the categorical axes. In `{ggplot2}`, it’s incredibly straightforward. 

## The Waterfall Plot

A Waterfall Plot aligns every patient on the X-axis (sorted by their tumor shrinkage) and plots a vertical bar representing their maximum percentage change from baseline in the sum of target lesion diameters.

```{r}
#| label: waterfall
#| message: false
library(dplyr)
library(ggplot2)
library(forcats)

# Simulated ADTR (Tumor Response) data
set.seed(42)
adtr_bor <- data.frame(
  USUBJID = paste0("S-", 1:40),
  PCHG = rnorm(40, mean = -10, sd = 30), # % Change from Baseline
  BOR = sample(c("CR", "PR", "SD", "PD"), 40, replace = TRUE, prob = c(0.1, 0.3, 0.4, 0.2))
)

# Cap growth at +100% just for visualization scaling (Standard practice)
adtr_bor <- adtr_bor %>%
  mutate(PCHG_CAP = if_else(PCHG > 100, 100, PCHG))

# The key step for Waterfall: Reordering the X-axis by the Y-value
# We use forcats::fct_reorder to sort USUBJID factor levels by PCHG_CAP (descending)
ggplot(adtr_bor, aes(x = fct_reorder(USUBJID, -PCHG_CAP), y = PCHG_CAP, fill = BOR)) +
  
  geom_bar(stat = "identity", width = 0.8) +
  
  # Standard RECIST reference lines 
  # PR boundary: -30%
  # PD boundary: +20%
  geom_hline(yintercept = -30, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 20, linetype = "dashed", color = "black") +
  
  # Axis modifications
  scale_y_continuous(breaks = seq(-100, 100, 20), limits = c(-100, 100)) +
  scale_fill_manual(values = c("CR" = "#1A85FF", "PR" = "#D41159", "SD" = "#FFC20A", "PD" = "#000000")) +
  
  theme_minimal() +
  theme(
    axis.text.x = element_blank(), # Hide individual subject IDs
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  labs(
    title = "Best Overall Response (Waterfall Plot)",
    x = "Individual Subjects",
    y = "Maximum % Change in Sum of Lesions from Baseline",
    fill = "Best Response"
  )
```

The trick here is `fct_reorder()`. In SAS, sorting a categorical axis by a numeric value requires creating an artificial `order` variable, sorting the dataset, and enforcing X-axis tick ordering. R handles it as an aesthetic mapping.

## The Spider Plot

A Spider Plot draws a line for each individual subject, tracking their tumor percent change across study visits.

```{r}
#| label: spider-plot
#| message: false

# Simulated longitudinal ADTR data
adtr_long <- data.frame(
  USUBJID = rep(paste0("S-", 1:10), each = 5),
  AVISITN = rep(seq(0, 24, by = 6), 10),
  PCHG = runif(50, -60, 40)
) %>%
  # Force baseline to 0
  mutate(PCHG = if_else(AVISITN == 0, 0, PCHG))

# Sorting subjects into "Responders" vs "Non-Responders" just for color coding
adtr_long <- adtr_long %>%
  group_by(USUBJID) %>%
  mutate(MIN_PCHG = min(PCHG)) %>%
  ungroup() %>%
  mutate(RespGroup = if_else(MIN_PCHG <= -30, "Objective Response", "No Objective Response"))


ggplot(adtr_long, aes(x = AVISITN, y = PCHG, group = USUBJID, color = RespGroup)) +
  geom_line(size = 0.8, alpha = 0.7) +
  geom_point(size = 2, alpha = 0.9) +
  
  # Standard RECIST reference lines
  geom_hline(yintercept = -30, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 20, linetype = "dashed", color = "black") +
  
  scale_color_manual(values = c("Objective Response" = "#1A85FF", "No Objective Response" = "#D41159")) +
  
  theme_bw() +
  labs(
    title = "Longitudinal Tumor Burden (Spider Plot)",
    x = "Study Visit (Weeks)",
    y = "% Change in Sum of Lesions from Baseline",
    color = "Response Status"
  )
```

With `ggplot2`, the `group = USUBJID` mapping instinctively knows to draw a unique, connected line for each subject across the discrete time points (`AVISITN`), ensuring zero "jump" lines between data points. 

With both SAS and R now capable of creating these visualizations, the critical difference is sheer speed of development. Once your `ggplot2` syntax is locked in, R produces these complex oncology metrics significantly faster and with less code complexity than traditional SAS macros.

Next up, we wrap up the TFL category by exploring **Data Listings** using Quarto!
