---
title: "Building OCCDS (Occurrence Data) ADaMs in R: Handling Adverse Events the R Way"
date: "2026-02-25"
author: "Harsh Verma"
categories: [ADaM, OCCDS, CDISC, admiral]
description: "How to merge ADSL into ADAE, derive treatment emergence flags, and handle occurrence data using {admiral}."
---

## Occurrence Data (OCCDS) Basics

OCCDS datasets like `ADAE` (Adverse Events), `ADCM` (Concomitant Medications), and `ADMH` (Medical History) deal with *occurrences*. These datasets usually merge the core domain (`AE`, `CM`, `MH`) with subject-level demographics (`ADSL`) to calculate flags based on timing relative to treatment.

The most critical calculation in `ADAE` is determining if an adverse event is **Treatment Emergent** (TEAE).

In SAS, this involves a data step comparing the AE start date (`ASTDT`) to the Treatment Start Date (`TRTSDT`) from `ADSL`.

```sas
/* SAS Code: TEAE Derivation */
data adae;
  merge sdtm_ae(in=a) adsl(in=b keep=USUBJID TRTSDT);
  by USUBJID;
  if a;

  TRTEMFL = '';
  if ASTDT >= TRTSDT and ASTDT ne . then TRTEMFL = 'Y';
run;
```

## Building ADAE with `{admiral}`

The `{admiral}` package streamlines OCCDS derivations. The workflow generally follows:
1. Load `AE` and `ADSL`.
2. Merge `ADSL` variables (like `TRTSDT`) into `AE`.
3. Derive partial dates/times (`ASTDT`, `AENDT`).
4. Derive population and occurrence flags (`TRTEMFL`, `CQ01NAM`).

Here is a simplified R script:

```{r}
#| eval: false
#| label: admiral-adae
#| message: false
library(dplyr)
library(admiral)
library(lubridate)

# Mock Data
ae <- data.frame(
  USUBJID = c("101", "101", "102"),
  AETERM = c("HEADACHE", "NAUSEA", "FATIGUE"),
  AESTDTC = c("2024-01-10", "2024-01-20", "2024-02-05")
)

adsl <- data.frame(
  USUBJID = c("101", "102"),
  TRTSDT = as.Date(c("2024-01-15", "2024-02-01"))
)

# 1. Merge ADSL
adae <- ae %>%
  derive_vars_merged(
    dataset_add = adsl,
    by_vars = exprs(USUBJID),
    new_vars = exprs(TRTSDT)
  ) %>%
  
  # 2. Convert ISO character dates to numeric R dates
  mutate(ASTDT = as.Date(AESTDTC)) %>%
  
  # 3. Derive Treatment Emergent Flag (TRTEMFL)
  derive_var_extreme_flag(
    by_vars = exprs(USUBJID),
    order = exprs(ASTDT),
    new_var = TRTEMFL,
    condition = ASTDT >= TRTSDT,
    mode = "first" # Just an example if we wanted the first event
  )

print(adae)
```

Wait, `derive_var_extreme_flag()` isn't quite right for a TEAE flag since TEAEs apply to *all* events that occur after treatment, not just the "extreme" first or last one.

Instead, `{admiral}` provides elegant logical evaluation via standard `{dplyr}` `mutate()`. Because `{admiral}` handles the complex merging (like the `derive_vars_merged` step above), the actual flag derivation is dead simple.

```{r}
#| eval: false
adae <- adae %>%
  mutate(
    TRTEMFL = if_else(ASTDT >= TRTSDT & !is.na(ASTDT), "Y", "N", missing = "N")
  )
```

## Handling First Occurrence Flags

Where `{admiral}` really shines is with **extreme flags**. In typical TFLs, we often report "Subjects with at least one AE." To do this, we need a flag (`AOCCIFL`) that indicates the *first occurrence* of any AE for a subject.

In SAS, this requires `PROC SORT` followed by `first.USUBJID`. 

In `{admiral}`:

```{r}
#| eval: false
adae_first <- adae %>%
  filter(TRTEMFL == "Y") %>%
  derive_var_extreme_flag(
    by_vars = exprs(USUBJID),
    order = exprs(ASTDT),
    new_var = AOCCIFL,
    mode = "first"
  )

adae <- adae %>%
  left_join(adae_first, by = names(adae)) %>%
  mutate(AOCCIFL = if_else(is.na(AOCCIFL), "", AOCCIFL))

print(adae)
```

This single function handles the sorting, grouping by `USUBJID`, checking the condition (`TRTEMFL == "Y"`), and flagging only the first record chronologically (`mode = "first"`).

## Summarizing AE Counts 

Once `ADAE` is built, we frequently need to count the number of AEs per subject and push that count back into `ADSL`.

```{r}
#| eval: false
# Add total AEs per subject back to ADSL
adsl_updated <- adsl %>%
  derive_var_merged_summary(
    dataset_add = adae,
    by_vars = exprs(USUBJID),
    new_var = AETOT,
    analysis_var = USUBJID,
    summary_fun = function(x) length(x)
  ) %>%
  mutate(AETOT = if_else(is.na(AETOT), 0, AETOT))

print(adsl_updated)
```

The `derive_var_merged_summary()` function eliminates the need for separate `PROC SQL` `count(*)` block and subsequent `merge`.

 OCCDS derivations in R highlight a massive shift from multi-step data processing (sort -> merge -> derive) to a single, continuous, highly readable text pipeline.

Next, we dive into the most complex of all ADaM structures: Basic Data Structure (BDS).
